Safe Memory Regions for Big Data Processing

### 2016-06-22 paper/type-inference.tex
type inference

 \item If the solver succeeds, it returns substitution functions $\substFn_\rho$ and
  $\substFn_\varphi$ for free region and predicate variables, respectively, introduced in
  step 1. We apply these substitutions to the elaborated program to produce the final program.

@@

%%    This is used to replace every FGJ type in the program
%%    compute a polymorphic region type template
%%     for every class and method from their FGJ types. The templates contain
%%     region variables ($\rho$) to denote unknown region annotations, and
%%     predicate variables ($\varphi$) to denote unknown constraints over
%%     such region annotations. Free region variables are generalized in these
%%     types (which are, hence, polymorphic).
%%  
%%    Second, we make use of the computed region type templates to elaborate
%% expressions by introducing region variables to denote unknown region
%% arguments in \C{new} expressions, method calls and function
%% applications.
%% While elaborating expressions, we also build a system of
%% constraints that capture well-formedness requirements and subtyping
%% relationships between type templates that must hold (as per the static
%% semantics in Fig.~\ref{fb-staticsem}) for the elaboration to be valid.
%% Third, we lift expression elaboration and constraint generation to
%% methods and classes. Finally, we solve the constraints by making use
%% of our fixpoint constraint solving algorithm \csolvestar, which
%% reduces the constraint solving problem to an abduction problem in a
%% Herbrand constraint system, and then relies on \csolve, our abduction
%% solver for that domain.

% Due to the presence of region-polymorphic
% recursion in \FB, constraints generated by the algorithm can be
% circular. More precisely, constraints generated can assume the form
% $\varphi \Leftrightarrow \phi \wedge F(\varphi)$, where $F$ is a
% non-idempotent substitution function for region variables in
% $\varphi$. In the constraint solving phase, the algorithm then relies
% on a fixpoint constraint solving algorithm called \csolvestar to solve
% the constraints and determine assignments for unknown region and
% predicate variables.

@@

\emph{Function Types}.
Since \FB{} is higher-order, fields of function type are allowed. The parameter instantiation step
instantiates function types as follows: the type of every parameter as well as the return value is
instantiated with fresh region identifiers, as described earlier, and finally these region identifiers
are generalized as formal region parameters of the function type.
For example, the function type $\C{List} \rightarrow \C{Object}$ is instantiated as
$\inang{\rho_0, \rho_1, \rho_2} \C{List} \inang{\rho_0,\rho_1} \rightarrow \C{Object}\inang{\rho_2}$.

@@

Recall that \emph{region identifiers} are used in four different roles.
A region identifier may serve as
(a) A \emph{formal region parameter} of a class or method, or
(b) a \emph{static region identifier} introduced by a \C{letregion} construct, or
(c) an  \emph{open transferable region identifier} introduced by an \C{open} construct, or
(d) a \emph{region variable}, introduced to represent an unknown actual region parameter
of a method invocation or object alloation,
which will be bound to a region identifier of one of the three
preceding kinds by the end of the type inference.

@@

\paragraph{Properties of generated constraints}
The set of generated constraints have two interesting properties that allow us to
solve the constraints efficiently.
%
The first property is that every region variable is forced to be unified with some
region constant by the constraints, except in one special case.
This follows because $\FB$ does not admit uninitialized variables.
(The only case where this does not hold is in the case of a recursive function
that calls itself in a non-terminating fashion: in this case, the function never
returns a value, and so the return-value can be typed as anything. In the sequel,
we assume that the return value of such a function is typed to be \C{unit}.)
% , and the type system uniquely determines the region parameters of every value
%
The second property is that the \emph{context region constraint} $\phictxt$
occurring on the antecedent of any validity constraint is a conjunction of
outlives-constraints of the form $\pi \outlives \pi_s$ where $\pi_s$
is a static region identifier and $\pi$ is a region constant (either a formal
region parameter, or static region identifier, or open transferable region identifier).
Furthermore, if $\phictxt$ includes any conjunct $\pi \outlives \pi_s$, then
it includes every conjunct $\pi_f \outlives \pi_s$ for every $\pi_f \in \predDeltaMap(\varphi)$.

@@

\textbf{TO DO:}
The entire algorithm is sound.
The second phase is complete as well: if there is any possible instantiation (of the
region variables and predicate variables) of the elaborated program produced by the first
phase that type checks, then the second phase will identify it.
Incompleteness exists only in a few choices made during the first phase (elaboration phase).

>>>>

 \item If the solver succeeds, it returns substitution functions $\substFn_\rho$ and
  $\substFn_\varphi$ for free region and predicate variables, respectively, introduced in
  step 1. We apply these substitutions to the elaborated program to produce the final program.
\end{enumerate}

@@

We later on establish the soundness of the type inference algorithm, and also establish
partial completeness results.

@@

\emph{Function Types}.
Since \FB{} is higher-order, fields of function type are allowed. The parameter instantiation step
instantiates function types as follows: the type of every parameter as well as the return value is
instantiated with fresh region identifiers, as described earlier, and finally these region identifiers
are generalized as formal region parameters of the function type.
For example, the function type $\C{List} \rightarrow \C{Object}$ is instantiated as
$\inang{\rho_0, \rho_1, \rho_2} \C{List} \inang{\rho_1,\rho_2} \rightarrow \C{Object}\inang{\rho_0}$.

@@

\emph{Region identifiers} are of two kinds: \emph{region constants} and \emph{region variables}.
A \emph{region variable} is introduced to represent an unknown actual region parameter
of a method invocation or object alloation, which will be bound to a region constant by the end of the
type inference.
A region constant may be 
(a) a \emph{formal region parameter} of a class or method, or
(b) a \emph{static region identifier} introduced by a \C{letregion} construct, or
(c) an  \emph{open transferable region identifier} introduced by an \C{open} construct.

@@

\paragraph{Properties of generated constraints}
The set of generated constraints have two interesting properties that allow us to
solve the constraints efficiently.
%
The first property is that every region variable is guaranteed to be unified with some
region constant by the constraints\footnote{
There is one special case where this does not hold: a recursive function
that calls itself in a non-terminating fashion. Such a function never
returns a value, and so the return-value can be typed as anything.
We assume that the return value of such a function is typed to be \C{unit},
which will not introduce any region variables at the call-site of such a function.}
This follows because $\FB$ does not admit uninitialized variables.
% , and the type system uniquely determines the region parameters of every value
%
The second property is that the \emph{context region constraint} $\phictxt$
occurring on the antecedent of any validity constraint is a conjunction of
outlives-constraints of the form $\pi \outlives \pi_s$ where $\pi_s$
is a static region identifier and $\pi$ is a region constant (either a formal
region parameter, or static region identifier, or open transferable region identifier).
Furthermore, if $\phictxt$ includes any conjunct $\pi \outlives \pi_s$, then
it includes every conjunct $\pi_f \outlives \pi_s$ for every $\pi_f \in \predDeltaMap(\varphi)$.

@@

The type inference algorithm is sound:

Furthermore, the second phase of the type inference algorithm is complete as well.
If there is any possible instantiation (of the region variables and predicate variables) of
the elaborated program produced by the first phase that type checks, then the second phase
will identify it.

Incompleteness exists only in a few choices made during the first phase (elaboration phase), as
explained below.

\begin{enumerate}
\item Our technique for determining the set of region parameters for a recursive class
incorporates a specific heuristic, namely that the recursive occurrence of the class has
the same parameters, in the same order, as the class itself. This technique will fail, for
example. if the program uses a recursive list type whose elements alternatively come
from two different regions.
% , $R_1$ and $R_2$.
% both of which are region parameters of the class.

\item Our technique for region parametrization also uses a heuristic in the case of
higher order programs. Consider a higher-order method that has a function $f$ as a
parameter. We have a choice in terms of where the region-parameters introduced for
$f$ are bound (i.e., quantified). We always bind these region-parameters within the type
of $f$ (essentially, requiring $f$ to be region-polymorphic). This may also be a potential
source of incompleteness in the type-inference.

The treatment of class fields with a function type also uses the same (incomplete)
heuristic.

Region annotations provided by the user can help the type-inference overcome these
sources of incompleteness.
